Baseline centering problem

### 2019-01-14 paper/icml_drau/framework.tex

\subsection{Framework}

@@

@@

\paragraph{Return and Discount Factor} For a continuous MDP the return is defined as \TODO{Cite}:
\begin{equation}
  \label{eq:continuous-return}
  R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt
\end{equation}

The natural discretization of this equation, if we want to define a discretize return $R_\delta$ for each $\deltat$ is:
\begin{equation}
  \label{eq:discretized-return}
  R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}r(s^\deltat_k, a^\deltat_k)\deltat
\end{equation}

We see that the natural definitions of $\gamma_\deltat$ and $r_\deltat$ such that the return defined in equation (\ref{eq:discretized-return}) corresponds to the discrete return for the MDP ${\cal M}_\deltat$ are:
\begin{align}
  \label{eq:def-gamma}
  \gamma_\deltat =& \gamma^\deltat\\
  \label{eq:def-reward}
  r_\deltat =& \deltat . r
\end{align}
% Thus, we see that the natural discretization of a continuous MDP ${\cal M} = \langle {\cal S}, {\cal A}, T, r, \gamma\rangle$ is ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_\deltat, r\deltat, \gamma^\deltat\rangle$.

@@

\paragraph{Time horizon} The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon. We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps. But if the duration of a step is $\deltat$, the time horizon in \emph{physical time} is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
% This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$. \TODO{A changer pour ne pas ecrire de chose fausse}
But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical} quantity independant of $\deltat$. With $\gamma_\deltat = \gamma^\deltat$, the time horizon in \emph{physical time} is:
\begin{equation}
  \label{eq:time-horizon}
  \frac{\deltat}{1-\gamma^\deltat} \rightarrow_{\deltat \rightarrow 0} - \frac{1}{\log \gamma} \approx \frac{1}{1-\gamma}
\end{equation}

@@

With the definition of $\gamma_\deltat$ and $r_\deltat$ in equations \eqref{eq:def-gamma} and \eqref{eq:def-reward}, the Value Function for a policy $\pi$ with a discretization $\deltat$ becomes:
\begin{equation}
  V^\pi_\deltat(s) = \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
\end{equation}
We have the following convergence theorem of the discretized value function to the continuous value function:
  \begin{theorem}
    We have, for all $s \in {\cal S}$:
    \begin{equation}
      \label{eq:conv-value}
      \lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)
    \end{equation}
  \end{theorem}

@@

If the MDP is following the equation (\ref{eq:diffusion}), the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
\begin{equation}
  \label{eq:hamilton-jacobi-bellman}
  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = (1 - \gamma) V^\pi(s)
\end{equation}

@@

Contrary to the Value function, the Action-Value function $Q$ is ill-defined for a continuous MDP. More precisely, for a continuous MDP, $Q^\pi(s, a) = V^\pi(s)$, so the continuous $Q$-function is independant of the actions. For a \emph{near-continuous} MDP, this means that the $Q$ is very close  to the $V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$ function \TODO{explain}. This can be stated formally.
\begin{theorem}
  For all $s, a$, we have
  \begin{equation}
    \label{eq:q-discr}
    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
  \end{equation}

>>>>

\VC{Many reinforcement learning environments are not intrinsically time
	discrete, but discretization of an underlying continuous time
	environment. For instance, many control environments, such as the
	Mujoco~\cite{ddpg} environments or OpenAI Gym classic~!CITE! control
	environments are discretization of continuous time control problems.
	For real world environments, the discretization timestep, $\deltat$ is
	most often imposed by the quality of sensors and actuators. For visual
	simulated environments, it often corresponds to the number of frames
	that the simulator generates per second. As $\deltat$ is not an
	intrinsic parameter of the environment, but a constraint imposed by
	computational ressources, we would expect that decreasing $\deltat$
	could  only improve the performance of reinforcement learning agents.
	To achieve this goal, reinforcement learning algorithms should be
	resilient to a change of $\deltat$, and should remain viable when
	$\deltat \rightarrow 0$. We are thus interested in the behavior of
	reinforcement learning algorithms in discretized environments, when the
	discretization timestep is small. We will refer to such environments as
	\emph{near continuous environments}.

	In what follows, a formalized view of near continuous environments is given,
	along with $\deltat$ dependant definitions of return, discount factor and
	value functions. With these definitions, each of these quantities is shown
	to converge to a well defined continous time limit. Finally, it is shown
	that the state-action value function collapses to the value function as
	$\deltat$ goes to $0$, and consequently that there is no Q-learning in
	continuous time.
}
\subsection{Framework}

@@

\VC{Consider ${\cal S} = \mathbb{R}^d$ a set of states, and ${\cal A}$ a set of
	actions. A familly of \emph{Markov decision processes} ${\cal
	M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
	\gamma_\deltat\rangle$ is \emph{near continous} if for any
	deterministic policy $\pi:{\cal S} \rightarrow {\cal A}$, trajectories
	of ${\cal M}_\deltat$ converge to stochastic processes that satisfy the
	Stochastic differential equation
	\begin{equation}
		\label{eq:diffusion}
		ds = F(s, a)dt + \Sigma(s, a)\dbt.
	\end{equation}
	Informally $F: {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
	the deterministic dynamic of the environment, while $\Sigma(s, a)\dbt$
	describes its stochasticity. Notably, $\Sigma = 0$ if the environment is
	deterministic, as in classical control theory, or in many usual simulated
	environments (Mujoco, Gym, ...).

	In what follows, ${\cal M}_\deltat$ is taken to be a direct discretization of
	Eq.~\ref{eq:diffusion} with discretization timestep $\deltat$. The
	corresponding transition function $T_\deltat(., s, a)$ is the distribution of
	$s'$ when starting from point $s$, and following Eq.~\ref{eq:diffusion} while
	maintaining $a$ constant on an interval of time $\deltat$.

	In this paper, we try to design RL methods for \emph{near continuous} MDP
	robust to the choice of $\deltat$. Our method is to analyse the different
	quantities we are looking at when $\deltat$ is changing, and especially when
	$\deltat \rightarrow 0$, and make sure that the quantities we are interested in
	and our algorithms do converge to meaningful limit quantities and algorithms.
	!OK?! The first formal statement we can make on discretized MDP with a
	time discretization $\deltat$ is that if $\deltat \rightarrow 0$, the
	distribution of trajectories do converge to the continuous trajectories.

	\begin{theorem}
		Assume that $\pi: {\cal S} \rightarrow {\cal A}$ is a continuous and
		deterministic policy. Let $(s_t)_{t\in\mathbb{R}_+}$ be the
		continuous process by following \eqref{eq:diffusion} and
		$(s_\deltat^k)$ be the discrete process obtained if the agent only
		get observations every $\deltat$ !explain better!.

		We define the time process $\tilde s_\deltat^t = s^{\lfloor
		\frac{t}{\deltat}\rfloor}_\deltat$ for
		$t\in\mathbb{R_+}$.!Explain!. Then:
		\begin{equation}
			\label{eq:conv-traj}
			(\tilde s_\deltat^t)_{t\in\mathbb{R}_+} 
			\xrightarrow[\deltat \rightarrow 0]{} (s_t)_{t\in\mathbb{R}_+}
		\end{equation}
		in distribution.
	\end{theorem}
}

@@

\VC{
	In what follows, $\deltat$-dependant definitions of the discount factor
	and the reward are given. These definitions fit the discrete case
	scenario when $\deltat=1$, and yields well defined, non trivial returns
	and value functions when $\deltat$ goes to $0$.
}

\paragraph{Return and Discount Factor}
For a continuous MDP and a continuous trajectory $\tau = (s_t, a_t)_t$,
the return is defined as~\cite{cont_rl}
\begin{equation}
	\label{eq:continuous-return}
	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt
\end{equation}

The natural discretization of this equation, if we want to define a discretized
return $R_\delta$ for each $\deltat$ is:
\begin{equation}
	\label{eq:discretized-return}
	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
	r(s^\deltat_k, a^\deltat_k)\deltat
\end{equation}

Natural definitions of $\gamma_\deltat$ and $r_\deltat$ such
that the return defined in equation (\ref{eq:discretized-return}) corresponds
to the discrete return for the MDP ${\cal M}_\deltat$ follow:
\begin{align}
	\label{eq:def-gamma}
	\gamma_\deltat =& \gamma^\deltat\\
	\label{eq:def-reward}
	r_\deltat =& \deltat . r.
\end{align}
% Thus, we see that the natural discretization of a continuous MDP ${\cal M} = \langle {\cal S}, {\cal A}, T, r, \gamma\rangle$ is ${\cal M}_\deltat = \langle {\cal S}, {\cal A}, T_\deltat, r\deltat, \gamma^\deltat\rangle$.

@@

\paragraph{Microscopic/Macroscopic time, Time horizon}
\VC{In near continuous environments, there are two notions of time,
	the number of observation received, or the number of actions taken,
	which we call the microscopic time, and the actual amount of time spent
	in the underlying continuous time environment, which we call the macroscopic
	time. The two are related through
	\begin{equation}
		\tau^\text{macro} = \tau^\text{micro} \deltat.
	\end{equation}

	The notion of time horizon, which is, informally, the amount of time on
	which the agent optimizes its return, is only meaningful in macroscopic
	time. Having a time horizon of $10$ steps means something very different
	if $\deltat = 0.1$ and if $\deltat = 0.0001$. A rule of thumb
	is that the time horizon of an agent with discount factor $\gamma$ is
	of order $\frac{1}{1 - \gamma}$. If $\gamma_\deltat$ was left constant
	as $\deltat$ goes to $0$, the corresponding macroscopic time horizon 
	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
	$\deltat$ goes to $0$.

	On the contrary, the definition of $\gamma_\deltat$ in
	Eq.~\eqref{eq:def-gamma} yields an approximately constant macroscopic
	time horizon when $\deltat$ goes to $0$. Indeed,
	\begin{equation}
		\label{eq:time-horizon}
		\frac{\deltat}{1-\gamma^\deltat}
		= - \frac{1}{\log \gamma} + \bigO(\deltat)
		\approx \frac{1}{1-\gamma}
	\end{equation}
}
The definition of $\gamma_\deltat$ in equation (\ref{eq:def-gamma}) is consistent with what we expect for the time horizon. We know \TODO{Cite?} that for a MDP with discount factor $\gamma$, the time horizon is $\frac{1}{1-\gamma}$ steps. But if the duration of a step is $\deltat$, the time horizon in \emph{physical time} is $\frac{\deltat}{1-\gamma}$, which goes to 0 if $\deltat \rightarrow 0$.
% This means that if $\gamma$ is not adapted to $\deltat$, the return $R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k}r(s^\deltat_k, a^\deltat_k)\deltat \rightarrow_{\deltat\rightarrow 0} r(s_0, a_0)$. \TODO{A changer pour ne pas ecrire de chose fausse}
But in practice, for example on a control problem, we do not care about the time horizon as a number of steps but as a \emph{physical} quantity independant of $\deltat$. With $\gamma_\deltat = \gamma^\deltat$, the time horizon in \emph{physical time} is:
\begin{equation}
  \label{eq:time-horizon}
  \frac{\deltat}{1-\gamma^\deltat} \rightarrow_{\deltat \rightarrow 0} - \frac{1}{\log \gamma} \approx \frac{1}{1-\gamma}
\end{equation}

@@

The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
following definition of the continuous value function
\begin{equation}
  V^\pi(s) = \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
\end{equation}
The corresponding discretization for time discretization $\deltat$ is
\begin{equation}
  V^\pi_\deltat(s) = \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
\end{equation}
which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
$r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
We have the following convergence theorem of the discretized value function to the continuous value function:
  \begin{theorem}
    We have, for all $s \in {\cal S}$:
    \begin{equation}
      \label{eq:conv-value}
      \lim_{\deltat \rightarrow 0} V^\pi_\deltat(s) = V^\pi(s)
    \end{equation}
  \end{theorem}

@@

If the MDP is following the equation (\ref{eq:diffusion}), the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:
\begin{equation}
  \label{eq:hamilton-jacobi-bellman}
  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = - \log(\gamma) V^\pi(s)
\end{equation}

@@

Contrary to the Value function, the Action-Value function $Q$ is ill-defined
for a continuous MDP. More precisely, for a continuous MDP, $Q^\pi(s, a) =
V^\pi(s)$, so the continuous $Q$-function is independant of the actions. For a
\emph{near-continuous} MDP, this means that the $Q$-function is very close  to the
$V$-function.\TODO{Say more ?} Moreover, this means that the discretized $Q$
function \TODO{explain}. This can be stated formally.
\begin{theorem}
  For all $s, a$, we have
  \begin{equation}
    \label{eq:q-discr}
    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
  \end{equation}

### 2019-01-14 paper/tangotd.tex

Building a universal policy. One may make the policy depend on a goal
representation $\omega$ as an additional input. One may simultaneously
compute $Z(s_1,s_2,\omega)$, namely the $Z$ function when following the
policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
policy parameters $\pi(a|s_1,\omega)$ such that following
$\pi(|\Omega(s_2))$ indeed goes towards $s_2$, namely, the policy
gradient for every $s_2$ is
\begin{equation}
\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
(z(s',s_2,\Omega(s_2))-z(s,s_2,\Omega(s_2)))
\end{equation}
[TODO check this is indeed the policy gradient for the time spent at
$s_2$ if following policy $\pi(|\Omega(s_2))$]. Advantage: diversity of
policies should appear by itself, no need to manually force options to be
distinct. Depends on a chosen distribution of goals $s_2$; we could learn
to go to uniformly distributed states, or create a goal at points where
we observe a high reward value (so learning to go to all states with a
reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
principle, but may be ambitious (or use a summary/auto-encoded value of
$s$). TODO check:
this is with a reward a Dirac at $s_2$, check there's no technical
problem wrt the reference distribution $\rho(s_2)$, as we use $z$ not $Z$
in the policy gradient and $z$ is the density of $Z$ wrt $\rho$ but I
think it's correct. Actually, for goal representations, a goal is not
necessarily a state itself, should be a feature of states, or some
representation of a reward function, so we should figure out the way to
give a meaning to $\Omega(R)$ for an arbitrary reward $R$. The
representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
thanks to linearity wrt $R$. This could even help with exploration:
setting temporary goals towards already-seen past states solves the "pure
random walk" problem by introducing temporary momentum; one can even
increase temperature a bit on $\omega$, by using slightly larger/more
variant values of $\omega$ as the goal.

>>>>

Building a universal policy. One may make the policy depend on a goal
representation $\omega$ as an additional input. One may simultaneously
compute $Z(s_1,s_2,\omega)$, namely the $Z$ function when following the
policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
policy parameters $\pi(a|s_1,\omega)$ such that following
$\pi(|\Omega(s_2))$ indeed goes towards $s_2$, namely, the policy
gradient for every $s_2$ is
\begin{equation}
\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
(z(s',s_2,\Omega(s_2))-z(s,s_2,\Omega(s_2)))
\end{equation}
This equation is used to train the policy parameters; $\Omega(s_2)$ is
trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2,\omega)$
(because the best way to reach $s_2$ is to select the $\omega$ with
maximal $z(s,s_2,\omega)$ at each step; TODO: from which $s$? TODO: here
I'm confusing constant action plans and time-variable action plans).
[TODO check this is indeed the policy gradient for the time spent at
$s_2$ if following policy $\pi(\Omega(s_2))$]. Advantage: diversity of
policies should appear by itself, no need to manually force options to be
distinct. Depends on a chosen distribution of goals $s_2$; we could learn
to go to uniformly distributed states, or create a goal at points where
we observe a high reward value (so learning to go to all states with a
reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
principle, but may be ambitious (or use a summary/auto-encoded value of
$s$). TODO check:
this is with a reward a Dirac at $s_2$, check there's no technical
problem wrt the reference distribution $\rho(s_2)$, as we use $z$ not $Z$
in the policy gradient and $z$ is the density of $Z$ wrt $\rho$ but I
think it's correct. Actually, for goal representations, a goal is not
necessarily a state itself, should be a feature of states, or some
representation of a reward function, so we should figure out the way to
give a meaning to $\Omega(R)$ for an arbitrary reward $R$. The
representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
thanks to linearity wrt $R$. This could even help with exploration:
setting temporary goals towards already-seen past states solves the "pure
random walk" problem by introducing temporary momentum; one can even
increase temperature a bit on $\omega$, by using slightly larger/more
variant values of $\omega$ as the goal.
